---
title: "Exponential distribution investigation"
author: "Srijan Nayak"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load_libraries}
library(ggplot2)
library(latex2exp)
```

## Overview

According to the *central limit theorem* (CLT), the distribution of sample means
approximates a normal distribution as the sample size increases, regardless of
the population distribution. This document aims to show the same for simulated
sample means of an exponential distribution with rate 0.2.

## Simulations

```{r random_seed}
# set the seed for random simulations
set.seed(42)
```

```{r sample_generation}
lambda <- 0.2
# get a sample of 1000 exponentials with lambda = 0.2
exp_sample <- rexp(1000, lambda)
# get 1000 means of 40 random exponentials with lambda = 0.2 each
means_sample <- apply(matrix(rexp(1000 * 40, lambda), ncol = 40), 1, mean)
```

A sample of 1000 random exponentials and a sample of 1000 means of 40 random
exponentials each are taken with $\lambda = 0.2$.

## Theoretical mean vs. sample mean

```{r mean_comparison, fig.height=3}
theoretical_mean <- 1 / lambda
sample_mean <- mean(means_sample)
# plot distribution of sample means with markers on theoretical and sample mean
ggplot() +
  geom_histogram(aes(x = means_sample), bins = 15) +
  geom_vline(aes(xintercept = theoretical_mean, color = "Theoretical mean"), size = 1) +
  geom_vline(aes(xintercept = sample_mean, color = "Sample mean"), size = 1) +
  scale_color_manual(name = "", values = c("#53DFA2", "#DF5390"), guide = "legend") +
  labs(title = "Distribution of sample means", x = TeX("$\\textit{\\bar{x}}$"))
```
As seen in the figure, the theoretical mean and the mean of the sample means are
very close. The theoretical mean, given by $\frac{1}{\lambda}$ is
`r theoretical_mean`, while the sample mean is `r sample_mean`.

## Theoretical variance vs. sample variance

The theoretical variance of the sample means is given by the square of the
standard error of the sampled exponential distribution, i.e.
$$
\textit{Variance of sample means}
= \textit{Standard error}^2
= \left( \frac{\frac{1}{\lambda}} {\sqrt[2]{n}} \right)^2
= \frac{1}{\lambda^2n}
$$

```{r variance_comparison}
# calculate the theoretical and sample variance of the sample means
theoretical_variance <- 1 / (lambda^2 * 40)
sample_variance <- sd(means_sample)^2
```

Using the above formula, theoretical variance is `r theoretical_variance`, while
the variance of the sample means is `r sample_variance`, which are again
reasonably close to each other.

## Distributions

```{r sample_distribution, fig.height=3}
ggplot() +
  geom_histogram(aes(x = exp_sample), bins = 15) +
  labs(title = "Distribution of exponential sample", x = TeX("$\\textit{x}$"))
```

The histogram of the 1000 random exponentials, as expected, looks like an
exponential distribution.

```{r sample_means_distribution, fig.height=3}
ggplot() +
  geom_histogram(aes(x = means_sample), bins = 15) +
  labs(title = "Distribution of sample means", x = TeX("$\\textit{\\bar{x}$"))
```

But the distribution of 1000 sample means, of 40 random exponentials each, looks
like a bell. Thus the sample means of 1000 random exponentials follow a normal
distribution as given by the central limit theorem.
